---
title: "Data Cleaning Final Project"
author: "Daniel Morehouse"
format: 
  html:
    embed-resources: true
    code-tools: true
    toc: true
editor: source
execute: 
  echo: false
  warning: false
---

```{r}
#| label: library-setup
#| include: false
# Installing packages essential for the project.

library(tidyverse)
library(dplyr)
library(readxl)
library(openxlsx)
```

This data set comes from my independent Marine Biology & Oceanography Research (MBOR) project that I conducted in Fall 2024 at the Bermuda Institute of Ocean Sciences. On 3 separate collection days, I incubated sponge fragments of 2 different species, one with an active chemical defense (T. ignis) and the other with no defense mechanism (D. etheria). Over the collection period, I wanted to determine if there was a change in thermal stress response between the 2 species. The data that will be used for this project comes from the first day of collection, 11/15/2024.

```{r}
#| label: read-data-excel

incubations <- read_excel("raw-data/MBORIncubations.xlsx", na = "N/A")
```

Through this data set, I wanted to answer 2 research questions:

- How does the surface area differ between the two experimental species?
- How does the respiration rate per hour differ between the ambient and heated treatments for each species?

In order to achieve these goals, I need to modify the data set to make it easier to work with. I'll be cleaning this data set in 3 ways to answer my 2 research questions:

1. Renaming variables in order to manipulate them into code (including modifying variables with mutate)
2. Selecting only the variables needed for data visualizations
3. Reformatting correct oxygen hour & surface area as variable type doubles rather than character

```{r}
#| label: renaming-variables
# Renamed variables for easy input into code, selected variables needed for data visualization

modified_incubations <- incubations |>
  rename(surface_area = `Surface Area (cm²)`) |>
  rename(oxy_hour = `Correct Oxy Hour (µmol/h)`) |>
  rename(species = `Sample ID`) |>
  rename(temp_treatment = Treatment) |>
  select(species, temp_treatment, surface_area, oxy_hour)
```

```{r}
#| label: change-variable-types
#| eval: false
# Reading the y-axis variables for both figures as double rather than character, using mutate function

modified_incubations |>
  mutate(surface_area = as.double(surface_area)) |>
  mutate(oxy_hour = as.double(oxy_hour))
```

```{r}
#| label: removing-blanks
# Removing the "cycle blanks" from the data set, only needed as controls to ensure the environments were stable
# Also modifying the species variable with a separate function to make it easier to plot

clean_incubations <- modified_incubations |>
  slice(-c(10, 20, 28, 36)) |>
  mutate(species_name = gsub("[ 0-9]", "", species))
```

```{r}
#| label: reading-in-clean-data
# Using write.xlsx to create a new sheet with my updated data set

write.xlsx(clean_incubations, file = "clean-data/clean_incubations.xlsx")
```

Now that the data is thoroughly cleaned and read in, I can move forward and create the desired figures to hopefully answer my research questions. 

```{r}
#| label: surface-area-figure

clean_incubations |> 
  ggplot(aes(y = species_name, 
             x = surface_area, 
             color = temp_treatment))+
  geom_boxplot() +
  labs(x = "Surface Area (cm^2)",
       y = "Sponge Species",
       color = "Temperature Treatment",
       title = "Difference in Surface Area Between Study Species") +
  theme_light() + 
  scale_color_manual(values = c("Ambient" = "Blue", "Heated" = "Red"))
```

This figure shows that the fragments of D. etheria were greater in size than the fragments of T. ignis - which corresponds with my experience from sampling. This is an important variable to consider when evaluating respiration rate as a greater surface area could impact the amount of gas exchange occurring.

```{r}
#| label: respiration-rate-figure

clean_incubations |> 
  ggplot(aes(y = species_name, 
             x = oxy_hour, 
             color = temp_treatment))+
  geom_boxplot() +
  labs(x = "Correct Oxygen Hour (µmol/h)",
       y = "Sponge Species",
       color = "Temperature Treatment",
       title = "Difference in Respiration Rate Between Study Species") +
  theme_light() + 
  scale_color_manual(values = c("Ambient" = "Blue", "Heated" = "Red"))
```

This figure shows that for both species, fragments in the heated treatment had a greater rate of respiration. Between species, the species with an active defense mechanism (T. ignis) had a greater rate of respiration in the heated treatment. It would require more data to evaluate whether or not this is a sign of heat-induced stress or productivity, but it is a great first step on the path to discovering a future bio-identifier in environments of temperature vulnerability.